Follow-up Samples

### 2017-04-14 submission/manuscript.tex

$2$ Department of Internal Medicine, University of Michigan, Ann Arbor, MI

$3$ Department of Family Medicine and Community Medicine, Penn State Hershey Medical Center, Hershey, PA

@@

\textbf{Background:} Colorectal cancer (CRC) is a worldwide health
problem and research suggests a correlation between the fecal bacterial
microbiome and CRC. Despite this, very little is known about what
happens to the microbiome after treatment for an adenoma or carcinoma.
This study tested the hypothesis that treatment for adenoma or carcinoma
results in changes towards a normal bacterial community. Specifically,
we tried to identify components within the community that were different
before and after treatment of adenoma, advanced adenoma, and carcinoma.

\textbf{Results:} There was a larger change in the bacterial community
in response to treatment for carcinoma versus adenoma (P-value
\textless{} 0.05) but not carcinoma versus advanced adenoma (P-value
\textgreater{} 0.05). There was a trend for increasingly less community
similarity, between samples pre- and post-treatment from adenoma to
advanced adenoma to carcinoma. Despite this, no difference was found in
the relative abundance of any specific OTU before and after treatment
for adenoma, advanced adenoma, or carcinoma groups (P-value
\textgreater{} 0.05). Using Random Forest models to assess whether
changes in post-treatment samples were towards a normal community, only
those with carcinoma had a significant decrease in positive probability
(P-value \textless{} 0.05); providing further evidence that treatment
has the greatest effect in those with carcinoma. The adenoma model used
a total of 62 OTUs, the SRN model used a total of 61 OTUs, and the
carcinoma model used a total of 59 OTUs. A total of 26 OTUs were common
to all three models with many classifying to commensal bacteria (e.g.
\emph{Lachnospiraceae}, \emph{Bacteroides}, \emph{Anaerostipes},
\emph{Blautia}, and \emph{Dorea}). Both chemotherapy and radiation did
not provide any additional changes to the bacterial community in those
treated for carcinoma (P-value \textgreater{} 0.05).

\textbf{Conclusions:} Our data partially supports the hypothesis that
the microbiome changes after treatment towards a normal community.
Individuals with carcinoma had more drastic differences to the overall
community than those with adenoma. Common OTUs to all models were
overwhelmingly from commensal bacteria, suggesting that these bacteria
may be important in initial polyp formation, development of advanced
adenoma, and transition to carcinoma.

@@

bacterial microbiome; colorectal cancer; polyps; FIT; post-surgery; risk
factors

@@

Colorectal cancer (CRC) is currently the third most common cause of
cancer deaths {[}1,2{]}. The rate of disease mortality has seen a
significant decrease, thanks mainly to improvements in screening
{[}1{]}. However, despite this improvement there are still approximately
50,000 deaths from the disease per year {[}2{]}. Some of these deaths
are due to disease recurrence, with current estimates that 20-30\% of
those undergoing resection having a CRC recurrence {[}3{]}. This is
important since approximately 30-50\% of those who have a recurrence
will die of CRC {[}4{]}. To reduce the burden on the healthcare system,
finding ways to better stratify those at highest risk of recurrence is
of great importance.

Recent studies in humans and mouse models have shown that altered
membership and structure of the gut microbiome correlate with CRC
pathogenesis {[}5--7{]}. Further, bacterial communities on the mucosa
are altered between normal and tumor tissue {[}8{]}. Collectively, these
studies provide a tantalizing link between our gut bacteria and CRC and
suggest the possibility of using biomarkers to diagnose disease. Indeed,
models created with microbiome data and clinical tests, such as Fecal
Immunoglobulin Test (FIT), result in good predictions of CRC {[}9,10{]}.
While these studies show how changed microbial communities or invasion
by inflammatory species can impact CRC progression {[}11{]}, they
provide very little information as to whether these communities change
and rebound towards normal after successful treatment of adenoma,
advanced adenoma, or carcinoma.

Providing an answer to this question is important because it has far
reaching implications on both how the bacterial community causes the
formation of more polyps {[}5,6{]} and the ability to use the microbiome
as a predictive screening tool {[}9,10{]}. Understanding polyp formation
and transition to advanced adenoma and then carcinoma is crucial to
being able to understand how to prevent CRC occurrence. Response of the
community to treatment is also equally important to predictive models
designed for screening purposes since an unresponsive community would
provide little additional information for important events, such as
recurrence {[}12{]}.

Using pre- (initial) and post- (follow up) treatment samples we tested
the hypothesis that treatment causes detectable changes towards a normal
microbiome in those with adenoma, advanced adenoma, and carincoma.
First, we assessed differences between samples pre- and post-treatment
in adenoma, advanced adenoma, or carcinoma using alpha or beta diversity
metrics. Second, we explored whether models built to classify adenoma,
advanced adenoma, or carcinoma versus normal were able to identify
specific community members that differed between pre- and post-treatment
samples. We also used these models to assess whether changes in the
community were toward a more normal microbiome. Finally, we assessed
both whether surgery for adenomas and advanced adenomas provided larger
community changes and whether chemotherapy or radiation provided
additional changes to the microbiome over surgical resection. This study
aims to provide evidence as to whether the altered CRC microbiome
persists or shifts back towards normal after such interventions.

@@

\textbf{\emph{The Bacterial Community:}} Within our 67-person cohort we
tested whether the microbiome in patients with adenoma (n = 22),
advanced adenoma (n = 19), or carcinoma (n = 26) had any broad
differences between pre- and post-treatment samples. We found that
carcinoma patients had a more dissimilar bacterial community between
these samples than those with adenoma (P-value \textless{} 0.001)
{[}Figure 1A{]}. Although no significant differences were observed
between advanced adenoma and carcinoma there was an increase in the
dissimilarity of pre- and post-treatment samples from adenoma (0.55 ±
0.21 (mean ± SD)) to advanced adenoma (0.65 ± 0.25) to carcinoma (0.78 ±
0.15) {[}Figure 1A{]}. The bacterial community structure pre-
post-treatment was visualized using NMDS for adenoma {[}Figure 1B{]}
(PERMANOVA \textgreater{} 0.05), advanced adenoma {[}Figure 1C{]}
(PERMANOVA \textgreater{} 0.05), and carcinoma {[}Figure 1D{]}
(PERMANOVA \textless{} 0.05). Interestingly, when pre- and
post-treatment samples were compared, regardless of whether they were
adenoma or carcinoma, there was no significant overall difference in
beta diversity (PERMANOVA \textgreater{} 0.05). There was no difference
between pre- and post-treatment samples when investigating alpha
diversity metrics for adenoma, advanced adenoma, or carcinoma {[}Table
S1{]}. Additionally, there was also no difference in the relative
abundance of any specific OTU between these samples for adenoma,
advanced adenoma, or carcinoma only {[}Figure S1{]}.

\textbf{\emph{Adenoma Model:}} Using normal and adenoma individuals from
a separate cohort we created an adenoma model to help assess whether
those with adenoma in this cohort changed towards normal. The range of
model AUC's from 100 runs of 20 repeated 10-fold cross-validation was
0.62 - 0.72 with the AUC of the model used for classification having an
AUC of 0.65 {[}Figure S2A{]}. There was a total of 62 OTUs in this model
with the vast majority classifying to bacteria typically thought of as
commensal {[}Figure S3A{]}. There was a significant difference between
the actual and predicted disease classification (P-value \textless{}
0.05). There was also no significant decrease in the positive
probability of adenoma between pre- and post-treatment samples (P-value
\textgreater{} 0.05) {[}Figure 2A{]}.

\textbf{\emph{Advanced Adenoma Model:}} Using normal and advanced
adenoma individuals from a separate cohort we created an advanced
adenoma model to help assess whether those with advanced adenoma in this
cohort changed towards normal. The range of model AUC's from 100 runs of
20 repeated 10-fold cross-validation was 0.68 - 0.77 with the AUC of the
model used for classification having an AUC of 0.73 {[}Figure S2B{]}.
There was a total of 61 OTUs in the advanced adenoma model {[}Figure
S3B{]}. Similar to the adenoma model the vast majority of OTUs
classified to bacteria typically thought of as commensal. Also similar
to the adenoma model there was a significant difference between the
actual and predicted disease classification (P-value \textless{} 0.05)
and no significant decrease in the positive probability of advanced
adenoma between pre- and post-treatment samples (P-value \textgreater{}
0.05) {[}Figure 2B{]}.

\textbf{\emph{Carcinoma Model:}} Using normal and carcinoma individuals
from a separate cohort we created a carcinoma model to help assess
whether those with carcinoma in this cohort changed towards normal. The
range of model AUC's from 100 runs of 20 repeated 10-fold
cross-validation was 0.84 - 0.9 with the AUC of the model used for
classification being 0.88 {[}Figure S2C{]}. Interestingly, the AUCs
improved from adenoma to advanced adenoma to carcinoma {[}Figure 2{]}.
There was a total of 59 OTUs in the carcinoma model {[}Figure S3C{]}.
Similar to the adenoma and advanced adenoma models the vast majority of
OTUs classified to bacteria typically thought of as commensal but OTUs
that also classified to \emph{Fusobacterium}, \emph{Porphyromonas}, and
\emph{Parvimonas} were also important for carcinoma classification
{[}Figure S3C{]}. Also, like the adenoma and advanced adenoma models
there was a significant difference between the actual and predicted
disease classification (P-value \textless{} 0.05). There was a
significant decrease in the positive probability of carcinoma between
pre- and post-treatment samples (P-value \textless{} 0.05) {[}Figure
2C{]}; suggesting that the carcinoma samples changed towards normal
after treatment, unlike either adenoma or advanced adenoma. The one
individual still positive for carcinoma after treatment had an increase
in carcinoma positive proability between their pre- and post-treatment
sample {[}Figure 2C{]}.

\textbf{\emph{Common OTUs:}} Next, to identify which OTUs were important
at all stages of disease, we identified common predictive OTUs within
the adenoma, advanced adenoma, and carcinoma models. When we compared
the three different models with each other there were a total of 26
common OTUs. Some of the most common taxonomic identifications belonged
to \emph{Bacteroides}, \emph{Blautia}, \emph{Anaerostipes},
\emph{Lachnospiraceae}, and \emph{Dorea}. The vast majority of the OTUs
that were common between these models had classifications to bacteria
typically thought of as commensal {[}Table S2{]}. The rank of these
OTUs, based on median mean decrease in accuracy (MDA), varied higly
depending on whether they were in the adenoma, advanced adenoma, or
carcinoma model {[}Figure 3{]}. As an example, \emph{Bacteroides}
(OTU85) was ranked 14\textsuperscript{th} in the carcinoma model but in
both the adenoma and advanced adenoma model ranked
35\textsuperscript{th} and 41\textsuperscript{st} respectively {[}Figure
3{]}.

\textbf{\emph{Treatment Affects on Community:}} After observing these
changes from treatment we assessed the possible confounding impact of
chemotherapy or radiation, in the carcinoma group, and surgical
resection, in the adenoma group, on the observed results. In the
carcinoma group neither chemotherapy nor radiation provided any
additional change from initial sample over surgery alone (P-value
\textgreater{} 0.05) {[}Table S3{]}. For the adenoma group there was a
single difference in observed OTUs (sobs) between those that received
surgerical resection and those that had regular polyp removal (P-value
\textless{} 0.05) {[}Table S4{]}. For the surgical resection comparison,
adenoma and advanced adenoma were combined due to the low number of
resection occurances in these two groups. Using a fisher exact test
there was no difference in the proportion of those receiving surgerical
resection between the adenoma and advanced adenoma groups (P-value
\textgreater{} 0.05). This data suggests that the microbiome changes
observed in the carcinoma group were mostly a result from surgerical
resection and not from chemotherapy or radiation.

@@

This study builds upon previous work from numerous labs that have
considered both how the bacterial community between those with and
without CRC differs and how it might be used as an early screening tool
{[}9,10,13--15{]}. Here we show that the microbiome changes towards
normal after treatment for carcinoma and that chemotherapy and radiation
did not provide any additional change. Although some of the important
OTUs classified to genera from bacteria considered the ``usual
suspects'' (e.g. \emph{Fusobacterium}, \emph{Porphyromonas}, and
\emph{Parvimonas}) many did not. The majority of important OTUs had
taxonomic classifications for resident gut microbes and were common for
the adenoma, advanced adenoma, and carcinoma models. This suggests that
members within the commensal community may be the first that change
during CRC pathogenesis. These subtle changes, in turn, could be the
first step in allowing more inflammatory bacteria to gain a foothold
within the colon {[}11{]}.

@@

Many of the common OTUs that we identified taxonomically classified to
potential butyrate producers (e.g. \emph{Clostridiales},
\emph{Roseburia}, and \emph{Anaerostipes}) {[}Table S2{]}. Other OTUs
classified to bacteria that are inhibited by polyphenols (e.g.
\emph{Bacteroides}). Both butyrate and polyphenols are thought to be
protective against cancer, in part by reducing inflammation {[}16{]}.
These protective compounds are derived from the breakdown of fiber,
fruits, and vegetables by resident gut microbes. One example of this
potential diet-microbiome-inflammation-polyp axis is that
\emph{Bacteroides}, which was highly prevalent in our models, are known
to be increased in those with high non-meat based protein consumption
{[}17{]}. High protein consumption in general has been linked with an
increased CRC risk {[}18{]}. Conversely, \emph{Bacteroides} are
inhibited by polyphenols which are derived from fruits and vegetables
{[}19{]}. Our data fits with the hypothesis that the microbial
metabolites from breakdown products within our own diet could not only
help to shape the existing community but also have an effect on CRC risk
and disease progression. Within this context the commensal community may
be an important modifiable risk factor for monitoring and preventing CRC
recurrence after treatment.

@@

\textbf{\emph{Study Design and Patient Sampling:}} Sampling and design
have been previously reported in Baxter, et al {[}9{]}. Briefly, study
exclusion involved those who had already undergone surgery, radiation,
or chemotherapy, had colorectal cancer before a baseline fecal sample
could be obtained, had IBD, a known hereditary non-polyposis colorectal
cancer, or familial adenomatous polyposis. Samples used to build the
models for prediction were collected either prior to a colonoscopy or
between 1 - 2 weeks after. The bacterial community has been shown to
normalize back to a pre-colonoscopy community within this time period
{[}20{]}. Our training cohort consisted of a total of 423 individuals
{[}Table 1{]}. Our study cohort consisted of 67 individuals with an
initial sample as described and a follow up sample obtained between 188
- 546 days after treatment of lesion {[}Table 2{]}. This study was
approved by the University of Michigan Institutional Review Board. All
study participants provided informed consent and the study itself
conformed to the guidelines set out by the Helsinki Declaration.

\textbf{\emph{16S rRNA Gene Sequencing:}} Sequencing was completed as
described by Kozich, et al. {[}21{]}. DNA extraction used the 96-well
Soil DNA isolation kit (MO BIO Laboratories) and an epMotion 5075
automated pipetting system (Eppendorf). The V4 variable region was
amplified and the resulting product was split between three sequencing
runs with normal, adenoma, and carcinoma evenly represented on each run.
Each group was randomly assigned to avoid biases based on sample
collection location. The pre- and post-treatment samples were sequenced
on the same run.

\textbf{\emph{Sequence Processing:}} The mothur software package
(v1.37.5) was used to process the 16S rRNA gene sequences and has been
previously described {[}21{]}. The general workflow using mothur was:
Paired-end reads were first merged into contigs, quality filtered,
aligned to the SILVA database, screened for chimeras, classified with a
naive Bayesian classifier using the Ribosomal Database Project (RDP),
and clustered into Operational Taxonomic Units (OTUs) using a 97\%
similarity cutoff with an average neighbor clustering algorithm. The
number of sequences for each sample was rarefied to 10523 to minimize
uneven sampling.

\textbf{\emph{Model Building:}} The Random Forest {[}22{]} algorithm was
used to create the model used to create the three models used. The
adenoma model classified normal versus adenoma, advanced adenoma was
normal versus advanced adenoma, and carcinoma was normal versus
carcinoma. The toal number of individuals in this data set was 423
individuals. There were a total of 239 individuals in the adenoma model,
262 individuals in the advanced adenoma model, and 266 indivdiuals in
the carcinoma model {[}Table 1{]}. Each model was then applied to our
67-person cohort testing prediction of adenoma intial (adenoma n = 22)
versus adenoma follow up (adenoma n = 0), advanced adenoma initial
(advanced adenoma n = 19) versus advanced adenoma follow up (advanced
adenoma n = 0), carcinoma initial (carinoma n = 26) versus carcinoma
follow up (carcinoma n = 1).

@@

\textbf{\emph{Statistical Analysis:}} The R software package (v3.3.2)
was used for all statistical analysis. Comparisons between bacterial
community structure utilized PERMANOVA {[}23{]} in the vegan package
(v2.4.1). Comparisons between probabilities as well as overall OTU
differences between pre- and post-treatment samples utilized a paired
Wilcoxson ranked sum test. Where multiple comparison testing was
appropriate, a Benjamini-Hochberg (BH) correction was applied {[}24{]}
and a corrected P-value of less than 0.05 was considered significant.
Unless otherwise stated the P-values reported are those that were BH
corrected. Model rank importance was determined by obtining the median
MDA from the 100 20 repeated 10-fold cross validation and then ranking
from largest to smallest MDA.

@@

\textbf{Figure 1: General Differences between Adenoma, Advanced Adenoma,
and Carcinoma Groups After Treatment.} A) A significant difference was
found between the adenoma and carcinoma group for thetayc (P-value =
NULL). B) NMDS of the pre- and post-treatment samples for the adenoma
group. C) NMDS of the pre- and post-treatment samples for the advanced
adenoma group. D) NMDS of the pre- and post-treatment samples for the
carcinoma group.

@@

\textbf{Figure S2: ROC Curves of the Adenoma, Advanced Adenoma, and
Carcinoma Models.} A) Adenoma ROC curve: The light green shaded areas
represent the range of values of a 100 different 80/20 splits of the
test set data and the dark green line represents the model using 100\%
of the data set and what was used for subsequent classification. B)
Advanced Adenoma ROC curve: The light yellow shaded areas represent the
range of values of a 100 different 80/20 splits of the test set data and
the dark yellow line represents the model using 100\% of the data set
and what was used for subsequent classification. C) Carcinoma ROC curve:
The light red shaded areas represent the range of values of a 100
different 80/20 splits of the test set data and the dark red line
represents the model using 100\% of the data set and what was used for
subsequent classification.

>>>>

$2$ Department of Internal Medicine, University of Michigan, Ann Arbor, M.

$3$ Department of Family Medicine and Community Medicine, Penn State Hershey Medical Center, Hershey, P.

@@

\textbf{Background.} Colorectal cancer (CRC) is a worldwide health
problem. Despite growing evidence that members of the gut microbiota can
drive tumorigenesis, little is known about what happens to the
microbiome after treatment for an adenoma or carcinoma. This study
tested the hypothesis that treatment for adenoma or carcinoma alters the
abundance of bacterial populations associated with with disease to those
associated with a normal colon. We tested this hypothesis by sequencing
the 16S rRNA genes in the feces of 67 individuals before and after
treatment for adenoma (N=22), advanced adenoma (N=19), and carcinoma
(N=26).

\textbf{Results.} There were large changes the the bacterial communities
associated with treatment across the three groups. The communities from
patients with carcinomas changed significantly more than those with
adenoma following treatment (P-value=5.4e-05); however, there was not a
significant difference between those with advanced adenoma and those
with adenoma or carcinoma (P-value\textgreater{}0.05). Although
treatment brought about large intrapersonal changes, the change in the
abundance of individual OTUs to treatment was not consistent within
diagnosis groups (P-value \textgreater{} 0.05). Because the distribution
of OTUs across patients and diagnosis groups was patchy, we used the
Random Forest machine learning algorithm to identify groups of OTUs that
allowed us to successfully distinguish between pre- and post-treatment
samples for each of the diagnosis groups. However, across the three
models, there was little overlap between the OTUs that were indicative
of treatment. Next, we used a larger cohort that contained individuals
with normal colons and those with adenomas, advanced adenomas, and
carcinomas to determine whether individuals who underwent a treatment
were more likely to have OTUs associated with normal colons. We again
built Random Forest models and measured the change in the positive
probability of having one of the three diagnoses. Although we could
clearly differentiate pre- and post-treatment communities from the three
diagnosis groups, only those patients that initially had carcinomas
experienced a significant decrease in positive probability of having a
carcinoma (P-value \textless{} 0.05). Finally, tested whether the type
of treatment impacted the microbiota of those diagnosed with carcinomas
and were unable to detect any significant differences in characteristics
of these communities between individuals treated with chemotherapy and
those treated with chemotherapy and radiation (P-value \textgreater{}
0.05).

\textbf{Conclusions.} Further exploration of the relationship between
diagnosis, treatment, and the impact on the microbiome will yield
improvements in disease management.

@@

bacterial microbiome; colorectal cancer; polyps; FIT; post-surgery; risk
factor.

@@

Colorectal cancer (CRC) is the third most common cause of cancer deaths
in the United States {[}1,2{]}. Disease mortality has significantly
decreased, thanks to improvements in screening {[}1{]}. Despite this
improvement, there are still approximately 50,000 deaths from the
disease per year {[}2{]}. Current estimates indicate that 20-30\% of
those who undergo treatment will experience recurrence and 30-50\% of
those patients will die {[}3,4{]}. Identification of methods to assess
patients' risk of recurrence is of great importance to reduce mortality
and healthcare costs.

There is growing evidence that the gut microbiota is involved in the
progression of CRC. Mouse-based studies have identified populations of
\emph{Bacteroides fragilis}, \emph{Escherichia coli}, and
\emph{Fusobacterium nucleatum} that alter disease progression
{[}5--9{]}. Furthermore, studies that alter the structure of the
microbiota through the use of antibiotics or inoculation of germ free
mice with human feces has shown that varying community compositions can
result in varied tumor burden {[}10--12{]}. Collectively, these studies
hypothesize that the microbiota is altering the amount of inflammation
in the colon and with it the rate of tumorigenesis {[}13{]}.

Building on this evidence, several human-based studies have identified
unique signatures of colonic lesions {[}14--19{]}. One line of research
has identified community-level differences between those bacteria that
are found on and adjacent to colonic lesions and have supported a role
for \emph{Bacteroides fragilis}, \emph{Escherichia coli}, and
\emph{Fusobacterium nucleatum} in tumorigenesis {[}20--22{]}. Others
have proposed feces-based biomarkers that could be used to diagnose the
presence of colonic adenomas and carcinomas {[}23--25{]}. These studies
have associated \emph{Fusobacterium nucleatum} and other oral pathogens
with lesions. They have also noted that the loss of bacteria generally
thought to produce short chain fatty acids, which can suppress
inflammation, is associated with colonic lesions. Gut bacteria have a
role in tumorigenesis and there is promise that these populations may be
useful biomarkers for aiding in the early detection of disease.

Despite advances in understanding the role between the gut microbiota
and colonic tumorigenesis, we still do not understand how treatments
including resection, chemotherapy, and/or radiation affect the
composition of the gut microbiota. If the community drives tumorigenesis
then one would hypothesize that treatment to remove a lesion would
affect the microbiota and risk of recurrence. To test this hypothesis,
we addressed two related questions: Does treatment affect the colonic
microbiota in a predictable manner? If so, does the treatment alter the
community to more closely resemble that of individuals with normal
colons?

We answered these questions by sequencing the V4 region of 16S rRNA
genes amplified from fecal samples of individuals with adenoma, advanced
adenoma, and carcinomas pre- and post-treatment. We used both classical
community analysis to compare the alpha and beta-diversity of
communities pre- and post treatment. Next, we generated Random Forest
models to identify bacterial populations that were indicative of
treatment for each diagnosis group. Finally, we measured the predictive
probability of diagnosis models to determine whether treatment would
alter the probability of being diagnosed as having a normal colon or the
colon of the original diagnosis group. Ultimately, we found that
treatment alters the composition of the gut microbiota; however, we only
observed a shift from diseased to normal colon for those initially
diagnosed with carcinomas. Understanding how the community responds to
these treatments could be a valuable tool for identifying biomarkers to
quantify the risk of recurrence.

@@

\textbf{\emph{Treatment alters the bacterial community structure of
patients diagnosed with colonic lesions.}} Within our 67-person cohort
we tested whether the microbiome of patients with adenoma (N=22),
advanced adenoma (N=19), or carcinoma (N=26) had any broad differences
between pre- and post-treatment samples (Table 1). The structure of the
microbial communities of the pre- and post-treatment samples differed
{[}Figure 1A{]}. We found that the communities obtained pre- and
post-treament among the patients with carcinomas changed significantly
more than those patients with adenoma (P-value \textless{} 0.001). There
were no significant differences in the amount of change observed between
the patients with adenoma and advanced adenoma or between the patients
with advanced adenoma and carcinoma (P\textgreater{}0.05). Next, we
tested whether there was a consistent direction in the change in the
community structure between the pre and post-treatment samples for each
of the diagnosis groups {[}Figure 1BCD{]}. We only observed a consistent
shift in community structure for the patients with carcinoma when using
a PERMANOVA test (adenoma P-value0.999, advanced adenoma P-value0.945,
and carcinoma P-value0.005). Finally, we measured the number of observed
OTUs, Shannon evenness, and Shannon diversity in the pre and
post-treatment samples and did not observe a significant change for any
of the diagnosis groups (P-value\textgreater{}0.05) {[}Table S1{]}.

\textbf{\emph{The effects of treatment are not consistent across
treatment groups.}} We used two approaches to identify those bacterial
populations that change between the two samples for each diagnosis
group. First, we sought to identify individual OTUs that could account
for the change in overall community structure. However, using a paired
Wilcoxon test we were unable to identify any OTUs that were
significantly different in the pre and post-treatment groups {[}Figure
S1{]}. It is likely that high inter-individual variation and the patchy
distribution of OTUs across individuals limited the statistical power of
the test. To overcome these problems we developed Random Forest models
to identify collections of OTUs that would allow us to differentiate
between pre and post-treatment samples from each of the diagnosis
groups. Because of the relatively small number of subject in each group,
we restricted our models to only incorporate 10 OTUs. Despite this
restriction, the models performed well (adenoma AUC=0.69 - 0.92,
advanced adenoma AUC=0.8 - 1, carcinoma AUC=0.82 - 0.98). Interestingly,
the 10 OTUs that were used for each model had little overlap with each
other {[}Figure 2{]}. These results support the earlier community-wide
analysis where we observed that the treatment had an impact on the
overall community structure; however, the effect of treatment was not
consistent across patients and diagnosis groups.

\textbf{\emph{Post-treatment samples from patients with carcinoma
share}} Next, we determined whether treatment changed the microbiota in
a way that the post-treatment communities resembled that of patients
with normal colons. To test this, we used an expanded cohort of 423
individuals that were diagnosed under the same protocol as having normal
colons or colons with adenoma, advanced adenoma, or carcinoma (Table 2).
We then constructed Random Forest models to classify the pre and
post-treatment samples as having their original diagnosis or having a
normal colon. The models performed well (adenoma AUC=0.62 - 0.72,
advanced adenoma AUC=0.68 - 0.77, carcinoma AUC=0.84 - 0.9; Figure S2).
The OTUs that were most important in the models for differentiating
between patients with normal colons and those with adenomas or advanced
adenomas were dominated by populations commonly considered as commensals
(e.g. \emph{Faecalibacterium}, \emph{Lachnospiraceae},
\emph{Bacteroides}, \emph{Dorea}, \emph{Anaerostipes}, and
\emph{Roseburia}) {[}Figure S3{]}. Although many of these OTUs were also
included in the model differentiating between patients with normal
colons and those with carcinoma, this model also included OTUs
affiliated with populations that have previously been associated with
carcinoma (\emph{Fusobacterium}, \emph{Porphyromonas},
\emph{Parvimonas}) {[}Figure S3{]}. Finally, we applied these three
models to the pre and post-treatment samples for each diagnosis group
and quantified the change in the positive probability of the model. A
decrease in the positive probability would indicate that the colon more
closely resembled that of a patient with a normal colon. There was not a
significant change in the positive probability for the adenoma or
advanced adenoma groups {[}Figure 3{]}. The positive probability for the
pre and post-treatment samples from patients diagnosed with carcinoma
significantly decreased. In fact, only 6 of the 26 patients (23.08\%)
that were initially diagnosed with a carcinoma had a higher positive
probability after treatment. One of those was re-diagnosed with
carcinoma on the follow up visit. These results indicate that although
there are changes in the microbiota associated with treatment, those
experienced by the patients in the carcinoma group were the only ones
where the change was directed towards what would be found in a normal
colon

\textbf{\emph{Difficult to identify effects of specific treatments on
the change in the microbiota.}} The type of treatment that the patients
received varied across diagnosis groups. Those with adenomas and
advanced adenomas received surgical resection (adenoma, N=4; advanced
adenoma, N=4) or polyp removal during colonoscopy (adenoma, N=18;
advanced adenoma, N=15) and those with carcinomas received surgical
resection (N=12), surgical resection with chemotherapy (N=9), and
surgical resection with chemotherapy and radiation (N=5). We focused on
the patients with carcinoma and pooled those patients that received
chemotherapy with those that received chemotherapy and radiation to
improve our statistical power. We did not observe a significant
difference in the effect of these treatments on the number of observed
OTUs, Shannon diversity, or Shannon evenness (P-value \textgreater{}
0.05). Furthermore, there was not a significant difference in the effect
of the treatments on the amount of change in the community structure (P=
1). Finally, the change in the positive probability was not
significantly different between the two treatment groups (P=1). Due to
the relatively small number of samples in each treatment group, it is
difficult to make a definitive statement regarding the specific type of
treatment on the amount of change in the structure of the microbiota

@@

This study builds upon previous work from numerous labs that have
considered both how the bacterial community between those with and
without CRC differs and how it might be used as an early screening tool
{[}23--27{]}. Here we show that the microbiome changes towards normal
after treatment for carcinoma and that chemotherapy and radiation did
not provide any additional change. Although some of the important OTUs
classified to genera from bacteria considered the ``usual suspects''
(e.g. \emph{Fusobacterium}, \emph{Porphyromonas}, and \emph{Parvimonas})
many did not. The majority of important OTUs had taxonomic
classifications for resident gut microbes and were common for the
adenoma, advanced adenoma, and carcinoma models. This suggests that
members within the commensal community may be the first that change
during CRC pathogenesis. These subtle changes, in turn, could be the
first step in allowing more inflammatory bacteria to gain a foothold
within the colon {[}13{]}.

@@

Many of the common OTUs that we identified taxonomically classified to
potential butyrate producers (e.g. \emph{Clostridiales},
\emph{Roseburia}. and \emph{Anaerostipes}) {[}Table S2{]}. Other OTUs
classified to bacteria that are inhibited by polyphenols (e.g.
\emph{Bacteroides}). Both butyrate and polyphenols are thought to be
protective against cancer, in part by reducing inflammation {[}28{]}.
These protective compounds are derived from the breakdown of fiber,
fruits, and vegetables by resident gut microbes. One example of this
potential diet-microbiome-inflammation-polyp axis is that
\emph{Bacteroides}, which was highly prevalent in our models, are known
to be increased in those with high non-meat based protein consumption
{[}29{]}. High protein consumption in general has been linked with an
increased CRC risk {[}30{]}. Conversely, \emph{Bacteroides} are
inhibited by polyphenols which are derived from fruits and vegetables
{[}31{]}. Our data fits with the hypothesis that the microbial
metabolites from breakdown products within our own diet could not only
help to shape the existing community but also have an effect on CRC risk
and disease progression. Within this context the commensal community may
be an important modifiable risk factor for monitoring and preventing CRC
recurrence after treatment.

@@

\textbf{\emph{Study Design and Patient Sampling:}} Sampling and design
have been previously reported in Baxter, et al {[}23{]}. Briefly, study
exclusion involved those who had already undergone surgery, radiation,
or chemotherapy, had colorectal cancer before a baseline fecal sample
could be obtained, had IBD, a known hereditary non-polyposis colorectal
cancer, or familial adenomatous polyposis. Samples used to build the
models for prediction were collected either prior to a colonoscopy or
between 1 - 2 weeks after. The bacterial community has been shown to
normalize back to a pre-colonoscopy community within this time period
{[}32{]}. Our training cohort consisted of a total of 423 individuals
{[}Table 1{]}. Our study cohort consisted of 67 individuals with an
initial sample as described and a follow up sample obtained between 188
- 546 days after treatment of lesion {[}Table 2{]}. This study was
approved by the University of Michigan Institutional Review Board. All
study participants provided informed consent and the study itself
conformed to the guidelines set out by the Helsinki Declaration.

\textbf{\emph{16S rRNA Gene Sequencing:}} Sequencing was completed as
described by Kozich, et al. {[}33{]}. DNA extraction used the 96-well
Soil DNA isolation kit (MO BIO Laboratories) and an epMotion 5075
automated pipetting system (Eppendorf). The V4 variable region was
amplified and the resulting product was split between three sequencing
runs with normal, adenoma, and carcinoma evenly represented on each run.
Each group was randomly assigned to avoid biases based on sample
collection location. The pre- and post-treatment samples were sequenced
on the same run.

\textbf{\emph{Sequence Processing:}} The mothur software package
(v1.37.5) was used to process the 16S rRNA gene sequences and has been
previously described {[}33{]}. The general workflow using mothur was:
Paired-end reads were first merged into contigs, quality filtered,
aligned to the SILVA database, screened for chimeras, classified with a
naive Bayesian classifier using the Ribosomal Database Project (RDP),
and clustered into Operational Taxonomic Units (OTUs) using a 97\%
similarity cutoff with an average neighbor clustering algorithm. The
number of sequences for each sample was rarefied to 10523 to minimize
uneven sampling.

\textbf{\emph{Model Building:}} The Random Forest {[}34{]} algorithm was
used to create the model used to create the three models used. The
adenoma model classified normal versus adenoma, advanced adenoma was
normal versus advanced adenoma, and carcinoma was normal versus
carcinoma. The toal number of individuals in this data set was 423
individuals. There were a total of 239 individuals in the adenoma model,
262 individuals in the advanced adenoma model, and 266 indivdiuals in
the carcinoma model {[}Table 1{]}. Each model was then applied to our
67-person cohort testing prediction of adenoma intial (adenoma n = 22)
versus adenoma follow up (adenoma n = 0), advanced adenoma initial
(advanced adenoma n = 19) versus advanced adenoma follow up (advanced
adenoma n = 0), carcinoma initial (carinoma n = 26) versus carcinoma
follow up (carcinoma n = 1).

@@

\textbf{\emph{Statistical Analysis:}} The R software package (v3.3.2)
was used for all statistical analysis. Comparisons between bacterial
community structure utilized PERMANOVA {[}35{]} in the vegan package
(v2.4.1). Comparisons between probabilities as well as overall OTU
differences between pre- and post-treatment samples utilized a paired
Wilcoxson ranked sum test. Where multiple comparison testing was
appropriate, a Benjamini-Hochberg (BH) correction was applied {[}36{]}
and a corrected P-value of less than 0.05 was considered significant.
Unless otherwise stated the P-values reported are those that were BH
corrected. Model rank importance was determined by obtining the median
MDA from the 100 20 repeated 10-fold cross validation and then ranking
from largest to smallest MDA.

@@

**Figure 1: General Differences between Adenoma, Advanced Adenoma, and
Carcinoma Groups After Treatment.*. A) A significant difference was
found between the adenoma and carcinoma group for thetayc (P-value =
NULL). Solid black points represent the median value for each diagnosis
group. B) NMDS of the pre- and post-treatment samples for the adenoma
group. C) NMDS of the pre- and post-treatment samples for the advanced
adenoma group. D) NMDS of the pre- and post-treatment samples for the
carcinoma group.

@@

**Figure S2: ROC Curves of the Adenoma, Advanced Adenoma, and Carcinoma
Models.*. A) Adenoma ROC curve: The light green shaded areas represent
the range of values of a 100 different 80/20 splits of the test set data
and the dark green line represents the model using 100\% of the data set
and what was used for subsequent classification. B) Advanced Adenoma ROC
curve: The light yellow shaded areas represent the range of values of a
100 different 80/20 splits of the test set data and the dark yellow line
represents the model using 100\% of the data set and what was used for
subsequent classification. C) Carcinoma ROC curve: The light red shaded
areas represent the range of values of a 100 different 80/20 splits of
the test set data and the dark red line represents the model using 100\%
of the data set and what was used for subsequent classification.
